Waveform Inversion via Source Extension

### 2025-05-31 repro/papers/ssevelan/slides/slides.tex
near final

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\of}{\bar{f}}
\newcommand{\ou}{\bar{u}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\oJa}{\bar{J}_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}
\title[]{Waveform Inversion via Source Extension}
\author[]{William W. Symes \inst{1}}
\institute[]{\inst{1} Rice University}
\date{Workshop in honor of Jean Virieux \\June 2025}

@@

%\begin{frame}\frametitle{The Plan}
%\begin{itemize}
%\item What is ``extended modeling", and why bother
%\item FWI via surface source extension - examples
%\item Theory: why \& when it works
%\item Zoology of source extensions
%\end{itemize}
%\end{frame}

@@

For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F\bar{s} \approx d$. Update $m$ to drive
$\bar{w}(\bx,t;\bx_s) \rightarrow w(t)\delta(\bx-\bx_s)$ by minimizing
\[
  J_{\rm SSE}[m] = \sum_{\bx_s}\sum_{\bx,t} ||(\bx-\bx_s)|
  \bar{w}[m](\bx,t;\bx_s)|^2
 \]
\end{frame}

@@

provided that
\begin{itemize}
\item $d = F[m_*]s$,
\item RMS wavelength of $w$ (hence $d$) $\rightarrow 0$,
\end{itemize}
\[
  \nabla \nabla  J_{\rm SSE}[m_*] \rightarrow \sum_{\bx,\bp}
\]
\end{frame}

@@

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC1.pdf}
\end{center}
\end{frame}

\begin{frame}
HMC Salt/Carbonate Acoustic Model

Streamer acquisition $z_s = 7$ m, $\Delta x_s=37.5$ m, 315 shots, $z_r=15$ m, $\Delta x_r=12.5$ m, 475 recvs, max offset = $6$ km

Pulse for data generation = [2, 3, 6, 10] Hz trapezoidal bandpass filter

Free surface @ $z=0$, other boundaries absorbing, 2-8 staggered FD scheme

Regularization: penalize MS velocity gradient, weight (also $\alpha$) via several trials

Data contains diving waves, pre- \& post-critical reflections, multiple reflections, refracted multiple reflections, head waves, free surface multiples, ....

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC2.pdf}
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC3.pdf}
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\vspace{-0.7in}
\includegraphics[height=4.5in]{HMC4.pdf}
\end{center}
\end{frame}

\section{Theory: why \& when it works}

\begin{frame}
FWI vs. SSE for transmission (diving wave, crosswell) data:

$J_{\rm FWI}[m]$ and $J_{\rm SSE}[m]$ = regular functions of $m$, locally close to quadratic approximations.

$d$ consistent: $d=Ru[m,f]$, $\of = f = w(t)\delta(\bx-\bx_s)$, $A\of=0$ $\Rightarrow$

\begin{itemize}
\item  $J_{\rm FWI}[m]=J_{\rm SSE}[m]=0$
\vspace{0.5cm}
\item $J_{\rm FWI}[m + \delta m] \approx \mbox{const. }\omega^2 \|\delta m\|^2$ for some $\delta m$ (kinematic perturbations), $\omega$ = center frequency in source pulse $w(t)$
\vspace{0.5cm}
\item $J_{\rm SSE}[m+\delta m] \le \mbox{const. }\|\delta m\|^2$ independent of $\omega, \delta m$
\end{itemize}
\end{frame}

\begin{frame}
$\Rightarrow$ radius of convexity region:
\begin{itemize}
\item $J_{\rm FWI}: O(1/\omega)$, must initialize ``within a wavelength''
\item $J_{\rm SSE}: O(1)$, converges even with (rel.) poor initial guess, HF data
\end{itemize}

Method of proof: ray theory (``mocrolocal analysis'')

More detailed analysis: minimizing $J_{\rm SSE}$ equivalent to a form of transmission travel time tomography: {\color{blue} tomography without picking}

NB: for reflection data... ???
\end{frame}

\begin{frame}\frametitle{Late Breaking News}
Theory for special reflection case: layered ($\kappa(z)$) medium, dip filter to eliminate postcritical/diving wave energy

Explicit computation of $\nabla J_{\rm SSE}$ for 
\begin{itemize}
\item homogenous background 
\item data from single reflector in otherwise slowly varying $\kappa$, wavelet with central frequency $\omega$
\end{itemize}
$\Rightarrow$ $\nabla J_{\rm SSE} \rightarrow 0$ as $\omega \rightarrow \infty$

Same behavior as $\nabla J_{\rm FWI}$!!!

Upshot: for pure reflection data, SSE may not provide constructive velocity updates
\end{frame}

\begin{frame}
{\color{blue} {\bf Summary:}} Surface Source Extension

\begin{itemize}
\item overcomes cycle-skipping for transmission data, solid theoretical foundation
\item same computational complexity as FWI
\item open question: theory for reflections {\color{blue} (initial result not positive)}
\end{itemize}
%\end{frame}

%\begin{frame}
Other source extension methods (WRI, AWI, ...)

\begin{itemize}
\item Many promising examples 
\item Little known theoretically - connection to travel time tomography???
\end{itemize}
\end{frame}

\begin{frame}\frametitle{Model Extension}
Add degrees of freedom to $v$, $\kappa$, instead of source

Annihilator: flatten/focus gathers

{\color{blue}Extracts velocity updates from reflection and refraction events} (S. 86, Kern \& S. 94, S. GP 08, Shen \& S. GEO 08, Shen EAGE 12, Biondi \& Almomin 12, 14, Chauris \& Cocher GEO 17, Hou \& S. GEO 18, Barnier \& Biondi SEG 19)

Chief challenge: expense(model extension) $\gg$ expense(FWI) (Note: expense(source extension) $\approx$ expense(FWI))
\end{frame}

\begin{frame}\frametitle{Acknowledgements}
\begin{itemize}
\item Total E\&P USA
\item Sponsors of The Rice Inversion Project
\item Organizers of SS 1 (Recent Advances and Road Ahead) at SEG 2019
\end{itemize}
\end{frame}

\begin{frame}
Matched Source Waveform Inversion (Hua \& Symes 1993, 4, Huang \& Symes 2017)

\begin{itemize}
\item similar to AWI (source wavelet for each trace), w/o normalization
\item if
\begin{itemize}
\item transmission data (crosswell, diving wave)
\item unique ray paths
\end{itemize}
then mean square moment $\approx$ mean square travel time misfit: {\color{blue} tomography without picking}

\item relation with tomography fails ($\approx$ FWI) if multiple rays connect sources and receivers - contrast SSE
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}
Example 2: Deep Water Marmousi
\end{center}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmVt.pdf}\\
Modified Marmousi Model: 2 km water layer added to minimize non-reflected energy, also box-snaped target
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81}\\
Data: isotropic point source with 8 Hz Ricker wavelet, $z_r=0.04$ km, $x_r=$ 0.05 km to 9.15 km, $\Delta x_r=$ 0.05 km, $z_s=0.08$ km, $x_s=0.6$ km to $x_s=8.6$, $\Delta x_s$ = 0.1 km. Shots at $x_s=0.6$ km and $x_s=8.6$ km displayed.
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmV0.pdf}\\
Initial $v(z)$ model for both FWI and SSE
\end{center}
\end{frame}

\begin{frame}
\vspace{-0.1cm}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmFWI.pdf}\\
FWI estimate of $v$: 200 iterations LBFGS, frequency domain FD 5-8 Hz
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmSurf.pdf}\\
SSE estimate of $v$: 150 iterations LBFGS, frequency domain FD 5-8 Hz, fixed $\alpha$ set by trial
\end{center}
\end{frame}

\begin{frame}
%\vspace{0.07cm}
\begin{center}
\includegraphics[height=2in,width=4in]{Fig/marm2km/Marm2kmVt.pdf}\\
Modified Marmousi Model: 2 km water layer added to minimize non-reflected energy, also box-snaped target
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1Inv}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81Inv}\\
Data predicted from SSE inverted model: shots at $x_s=0.6$ km and $x_s=8.6$ km.
\end{center}
\end{frame}

\begin{frame}
\begin{center}
\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot1}\includegraphics[height=2in]{Fig/marm2km/Marm2kmShot81}\\
Target data:\\ shots at $x_s=0.6$ km and $x_s=8.6$ km.
\end{center}
\end{frame}

\begin{frame}
Wavefield Reconstruction Inversion: $\of(\bx,t;\bx_s)$ distributed throughout space-time

van Leeuwen and Herrmann 2013, 2016, Wang \& Yingst 2016

similar: space-time extension Huang \& S. 2018

\begin{itemize}
\item[+ ]many good examples, convincing numerical evidence
\item[+ ]3D field data - Wang \& Yingst 2016
\item[- ]basic form: store complete space-time (or space-frequency) fields (6D)
\end{itemize}
\end{frame}

\begin{frame}
Adaptive Waveform Inversion (Warner \& Guatsch, 2014, 2016): $\of(\bx,t;\bx_r,\bx_s) = w(t;\bx_r)\delta(\bx-\bx_s)$ localized but depends on receiver coordinate $\bx_r$ - one wavelet per data trace

similar: source-receiver extension Hua \& S. 1994, Huang \& S. 2017

\begin{itemize}
\item[+ ]additional storage: $\approx$ data volume (5D)
\item[+ ]minor mod of standard FWI workflow
\item[- ]source-receiver: loses advantage over FWI when sources, receivers connected by multiple ray paths (fails for lens model, above)
\end{itemize}
\end{frame}

\begin{frame}
Intermediate: source distributed over space-time surface
\begin{itemize}
\item const. t surface = volume source extension - Huang \& S. 2018
\item const z surface = {\color{blue} surface source extension (SSE):} 
\[
\of(\bx,t;x_s,y_s) = w(x,y; x_s,y_s)\delta(z-z_s)
\]
\end{itemize}
\begin{itemize}
\item[+ ]additional storage = O(data volume) (5D)
\item[+ ]no problem with multiple ray paths
\item[+ ]theory for transmission data (eg. diving wave): {\color{blue} tomography w/o picking}
\end{itemize} 
\end{frame}

>>>>

\AtBeginSection[]{
{\setbeamertemplate{logo}{}
 \begin{frame}<beamer>
    \frametitle{Agenda}
    \tableofcontents[currentsection]
  \end{frame}
}
}
% Images
% Macros
\newcommand{\peq}{\,+\hspace{-0.15cm}=}
\newcommand{\bm}{\bar{m}}
\newcommand{\dm}{\delta m}
\newcommand{\dom}{\delta \bar{m}}
\newcommand{\doma}{\delta \bar{m}_{\alpha}}
\newcommand{\om}{\bar{m}}
\newcommand{\bH}{{\bf H}}
\newcommand{\bR}{{\bf R}}
\newcommand{\bcF}{{\bar{\cal F}}}
\newcommand{\cF}{{\cal F}}
\newcommand{\oF}{\bar{F}}
\newcommand{\of}{\bar{f}}
\newcommand{\ou}{\bar{u}}
\newcommand{\odF}{\bar{F}^{\dagger}}
\newcommand{\oS}{\bar{S}}
\newcommand{\oM}{\bar{M}}
\newcommand{\oJa}{\bar{J}_{\alpha}}
\newcommand{\Na}{N_{\alpha}}
\newcommand{\dd}{\delta d}
\newcommand{\ddl}{\delta d_{\alpha}}

\title[]{Waveform Inversion via Source Extension}
\author[]{William W. Symes}
\institute[]{\inst{} Rice University}
\date{To Jean, for his many contributions and his leadership\\June 2025}

@@

@@

For any $m$, determine $\bar{w}=\bar{w}[m]$ so that $F[m]\bar{s} \approx d$. Update $m$ to drive
$\bar{w}(\bx,t;\bx_s) \rightarrow w(t)\delta(\bx-\bx_s)$ by minimizing
\[
  J_{\rm SSE}[m] = \sum_{\bx_s}\sum_{\bx,t} ||(\bx-\bx_s)|
  \bar{w}[m](\bx,t;\bx_s)|^2
 \]
\end{frame}

@@

provided that
\begin{itemize}
\item $d = F[m_*]s$,
\item $\Gamma \subset$ phase space over source surface, initial data
  for rays crossing receiver surface
\item $\tau[m](\bx_s,\bp)$ = time to receiver surface along ray taking
  off at $(\bx_s,\bp)$
\item RMS wavelength of $w$ (hence $d$) $\rightarrow 0$,
\end{itemize}
\[
  \mbox{Hess }  J_{\rm SSE}[m_*] \rightarrow \mbox{Hess } J_{\rm TT}[m_*]
\]
where $J_{\rm TT}[m] = \sum_{(\bx_s,p) \in \Gamma} (\nabla_p(\tau[m](\bx_s,\bp) -
\tau[m_*](\bx_s,\bp)))^2$

{\color{blue} SSE asymptotic to tomography-like limit}

\begin{frame}
Summary:

\begin{itemize}
\item for acoustic transmission data, several source extension methods
  perform ``tomography without picking'', avoid cycle-skipping,
  various constraints
\item no theoretical foundation for source extension with
  reflection inversion
\item beyond acoustics - elasticity? attenuation?
\end{itemize}

No shortage of open questions...
\end{frame}

@@
